---
title: "MATH2349 Semester 2, 2018"
author: "Richard Mailhot S3699115"
subtitle: Assignment 3
output:
  html_notebook: default
---


## Required packages 


```{r}
library(readr)
library(dplyr)
library(outliers)
```


## Executive Summary 


My data consists of three separate tables consisting of different NHL(National Hockey League) stats or information. I obtained this data from kaggle.com, with specific links given below. The game and game_stats tables only contain the teams that played using a key in the team info. I merged these so that I could have the actual team names in the data. When I merge all of these together I get over 30 columns, so I only selected the columns that I want to use in my final analysis.  I then create two additional numeric columns. Once I have my final data set in a tidy format, I check the data for erroneous entries. Next I check my numeric data for any outliers, before transforming my shots variable to normalize it.


## Data 

I use three separate data sets from kaggle.com.

They are 
team_info.csv from https://www.kaggle.com/martinellis/nhl-game-data#team_info.csv
game.csv from https://www.kaggle.com/martinellis/nhl-game-data#game.csv
game_teams_stats from https://www.kaggle.com/martinellis/nhl-game-data#game_teams_stats.csv

team_info contains information about each team. I use this data to get the abbreviation variable into my dataset. This variable allows me to easily see what team teh observation is about, without using a key.


games contains information about all games played from the 19th of January, 2013 to the 8th of June 2018 in the national hockey league. The variables I use from this data set are

* game_id - a unique id for each game
* season - which season is the game from
* type - playoff or regular season
* date_time - date of game played
* venue - what arena the game was played in

team_stats contains the team statistics from each of these games. I merge this with the games table, so that I have the date and type of game for each observation. The variables I use from this data set are

* game_id
* HoA - if the team was the home or away team
* won - whether this team won
* settled_in - if the game ended in regular time, overtime, or a shootout
* goals - goals scored by this team in this game
* shots - shots on goal
* powerPlayOpportunities- number of times on the power play
* powerPlayGoals- goals scored on the power play


```{r}

team_info <- read_csv("team_info.csv")
head(team_info)
games <- read_csv("game.csv")
head(games)
team_stats <- read.csv("game_teams_stats.csv")
head(team_stats)

team_stats_full <- team_stats %>% left_join(team_info, by = "team_id") %>% left_join(games, by = "game_id")
head(team_stats_full)
```

## Understand 

In this step i use the str() function to look at the types of data in each column of my data set. I notice that the date is imported as a character, so I change this to date format.

```{r}
str(team_stats_full)

team_stats_full$date_time <- as.Date(team_stats_full$date_time, format = "%d/%m/%Y")
class(team_stats_full$date_time)
head(team_stats_full)
```


##	Tidy & Manipulate Data I 

The full merged data is untidy. However the untidyness comes from variables I do not want in my final data set. For example the outcome variable tells us who won, and what part of the game it was settled in. I get rid of a lot of these columns when I select the columns I want for my final data set.


```{r}
 team_stats_full <- team_stats_full %>% select(c("game_id", "abbreviation", "HoA", "won", "settled_in", "head_coach", "goals", "shots", "powerPlayOpportunities", "powerPlayGoals", "season", "type", "date_time", "venue"))

head(team_stats_full)
```

##	Tidy & Manipulate Data II 

In this step I create two new variables. I create the powerPlayPercent variable, which tells us the percent of power plays that the team scored on, and the shooting percentage variable.


```{r}
team_stats_full <- team_stats_full %>% mutate(powerPlayPercent = 100*(powerPlayGoals/powerPlayOpportunities), shootingPercent = 100*(goals/shots))

head(team_stats_full)

```


##	Scan I 

In this step I scan the data for missing values. I find that the newly created variable powerPlayPercent has missing values. Upon looking into these missing values I find that these are created when a team had no power play opporunities in a game. This would create an erroneous calculation when creaeting the column above. I replace these NAs with 0s as that is representative of their power play percentage in these scenerios.

```{r}
colSums(is.na(team_stats_full))

team_stats_full$powerPlayPercent[is.na(team_stats_full$powerPlayPercent)] <- 0

```


##	Scan II

In this step I create a data frame containing only numeric entries of the full data set. I exclude game_id and season, as these are integer ids and not truly numeric values.

I boxplot each of these to see if they contain outliers, and also get the z-scores for each and see which are not in between -3 and 3.

```{r}

nums <- unlist(lapply(team_stats_full, is.numeric))

team_stats_numeric <- team_stats_full[ , nums] %>% select(-c("game_id", "season"))

boxplot(as.matrix(team_stats_numeric))
```

```{r}
goals_scores <- scores(team_stats_full$goals)
shots_scores <- scores(team_stats_full$shots)
PPPercent_scores <- scores(team_stats_full$powerPlayPercent)
PPOpp_scores <- scores(team_stats_full$powerPlayOpportunities)
PPGoal_scores <- scores(team_stats_full$powerPlayGoals)
shootingPercent_scores <- scores(team_stats_full$shootingPercent)

which(abs(goals_scores) > 3)
which(abs(shots_scores) > 3)
which(abs(PPPercent_scores) > 3)
which(abs(PPOpp_scores) > 3)
which(abs(PPGoal_scores) > 3)
which(abs(shootingPercent_scores) > 3)

```



##	Transform 

In this step I look at the shots variable. When looking at a histogram of the variable, we can see that it is slightly skewed left. To normalize this I take the square root of the variable. Upon viewing the new histogram we can see this does the job well.

```{r}
hist(team_stats_full$shots)

sqrt_shots <- team_stats_full$shots ^ .5

hist(sqrt_shots)
```




<br>
<br>
